Integration of Multiple Constraints in ACG

### 2014-02-17 lenls/post-proc.tex
Forked LENLS paper for post-proc
POST: Minor revisions, pangraphical models

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}

@@

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{gb4e}
\usepackage{bussproofs}
\usepackage{hyperref}
\usepackage{cancel}
\usepackage{pdfpages}
\usepackage{enumitem}

@@

A presupposed condition to actually using an ACG to describe discourse in a
large scope is to have a large scale ACG grammar in the first place. The work
presented here was motivated by this quest for a wide-coverage ACG grammar, as
seen from a point of view of language as a system of largely orthogonal
constraints. Encoding ancillary constraints in a type system can make the
system hard to read and understand. If multiple constraints written in the
style of \cite{pogodalla2012controlling} are to be enforced in the same
grammar, we advocate for extending the formalism to prevent the incidental
complexity that would otherwise emerge.

@@

In the above, we elaborate the types $NP$ and $N$ with features ($NEG{=}F$
and $NEG{=}T$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP_{NEG{=}T} \limp
N_{NEG{=}T} \limp N_{NEG{=}}\ldots$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

The intrinsic language is just the set of terms built on the node's
signature and having the node's distinguished type. The extrinsic
language is established by taking the extrinsic languages of its
predecessors, mapping them through lexicons and taking their
intersection, or just taking the node's intrinsic language if it has no
predecessors.

@@

The above characteristics motivated us to explore alternative interpretations
of G-ACGs. We will present one such alternative now, which exchanges the
language-algebraic point of view for a generative one.

@@

In the new paradigm, we interpret the nodes of the graph as terms and the
edges as statements that one term is mapped into another using a lexicon. This
leads us to the definition of the \emph{pangraphical}\footnote{As opposed to
  extrinsic languages, which are constrained only by their predecessors in the
  graph, pangraphical languages are constrained by the entire graph.} language
of a node $u$ in a G-ACG $\mathcal{G}$.

A term $t$ belongs to $\mathcal{P}_{\mathcal{G}}(u)$ whenever there
exists a labeling $T$ of the nodes of the graph such that:

\begin{itemize}
  \item $T_u = t$.
  \item For all $v \in V(G)$, $\vdash_{\Sigma_v} T_v : S_v$.
  \item For all $(v,w) \in E(G)$, $\mathcal{L}_{(v,w)}(T_v) = T_w$.
\end{itemize}

If we compare this new interpretation of G-ACGs to the former one, we find out
that in the case when the graph of the grammar is an arborescence, they are
actually equivalent. This means that in classical ACGs, where all the diagrams
are arborescences, the two metaphors (nodes as languages and nodes as terms)
can be, and are, used interchangeably. However, as we start to consider
non-arborescent graphs, we find, interestingly, that the two paradigms diverge
(i.e. $\exists \mathcal{G}, u.\ \mathcal{E}_{\mathcal{G}}(u) \neq
\mathcal{P}_{\mathcal{G}}(u)$).

@@

We start with the extraction constraints. \cite{pogodalla2012controlling}
describes three different constraints on extraction, all of them expressed
using the Constraint pattern (Figure~\ref{fig:single-cont}). We can take the
union of these three G-ACGs to get the G-ACG on Figure~\ref{fig:merged-conts}.
Next, we can incorporate knowledge from \cite{pogodalla2007generalizing}, by
splitting the $\mathcal{L}_{Syn}$ lexicon and introducing a new intermediate
signature (Figure~\ref{fig:conts-elab}). The new signature, $Syntax$, deals
purely with syntax without any issues of operator scope and its functions have
lower-order types than those in $Syn$. We will finally add our contribution,
the signatures $Neg$ and $Agr$ and the lexicons $\mathcal{L}_{Neg}$ and
$\mathcal{L}_{Agr}$ which implement the constraint on negation we discussed
before and some simple notion of agreement, respectively
(Figure~\ref{fig:final-synth}).

@@

We have decomposed our grammar into 9 signatures and 8 lexicons. Besides
making the grammar more readable and easily navigable, it also makes the
grammar more attractive from a computational point of view. The complexity of
parsing the highest abstract-most structures and verifying all the constraints
is decomposed into smaller subtasks. A string can be parsed into its
representation at the $Syntax$ level fairly easily (compared to parsing it at
the higher levels) since $Syntax$ employs types of lower orders than the more
abstract signatures. Finding the syntactic structure for a string is thus
feasible. We then proceed to verifying the syntactic constraints. Here we can
observe that the lexicons $\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ are mere
relabelings and therefore parsing (i.e. constraint verification) is trivial
(i.e. decidable). Now in order to determine the meaning of our sentence, we
have to continue parsing by finding an antecedent in $Syn$. This involves
higher-order types used to encode the various scope-bearing
operators. Finally, we can opt in to verify the high-level syntax-semantics
constraints which employ dependent types. However, the lexicons in this case
resemble relabelings and a semi-decidable procedure could be used to solve the
problem. Having the grammar decomposed like this opens it up for computational
processing by letting us engage in different tactics at different stages
instead of having to parse a single hidden structure which enforces
grammaticality w.r.t. every constraint. G-ACGs share these perks also with
classical ACGs but they let use this composition in another dimension, which
facilitates the writing of multiple constraints.

>>>>

\documentclass{llncs}
%
\usepackage{makeidx}

@@

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{gb4e}
\usepackage{bussproofs}
\usepackage{hyperref}
\usepackage{cancel}
\usepackage{pdfpages}
\usepackage{enumitem}

@@

A presupposed condition to actually using an ACG to describe discourse in a
large scope is to have a large scale ACG grammar in the first place. The work
presented here was motivated by this quest for a wide-coverage ACG grammar, as
seen from the point of view of language as a system of largely orthogonal
constraints. Encoding ancillary constraints in a type system can make the
system hard to read and understand. If multiple constraints written in the
style of \cite{pogodalla2012controlling} are to be enforced in the same
grammar, we advocate for extending the formalism to prevent the incidental
complexity that would otherwise emerge.

@@

In the above, we elaborate the types $NP$ and $N$ with features ($NEG{=}F$
and $NEG{=}T$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP_{NEG{=}T} \limp
N_{NEG{=}T} \limp N_{NEG{=}?}$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

The intrinsic language is just the set of terms built on the node's signature
and having the node's distinguished type. The extrinsic language is
established by taking the extrinsic languages of its predecessors, mapping
them through the lexicons and taking their intersection, or just taking the
node's intrinsic language if it has no predecessors.

@@

The above characteristics motivated us to explore alternative interpretations
of G-ACGs. We will present one such alternative now, which exchanges the
language-algebraic point of view for a term-relational one.

@@

In the new paradigm, we interpret the graph as a logical formula. The nodes
will be seen as variables whose domains are their intrinsic languages,
i.e. each node $v$ as a variable will range over the terms built from its
signature $\Sigma_v$ and having its distinguished type $S_v$. Edges $(v,w)$
will be statements saying that the image of the term $v$ through the lexicon
$\mathcal{L}_{(v,w)}$ is exactly the term $w$.

We can think of \emph{models} for this graphical propositional logic as
assignments of terms to variables, i.e. labelings of the nodes of the graph
with terms. A model $T$ is \emph{valid} when it satisfies the following
conditions:

\begin{itemize}
  \item For all $v \in V(G)$, $\vdash_{\Sigma_v} T_v : S_v$.
  \item For all $(v,w) \in E(G)$, $\mathcal{L}_{(v,w)}(T_v) = T_w$.
\end{itemize}

A term $t$ belongs to the \emph{pangraphical}\footnote{As opposed to extrinsic
  languages, which are constrained only by their predecessors in the graph,
  pangraphical languages are constrained by the entire graph, hence the name.}
language $\mathcal{P}_{\mathcal{G}}(u)$ if there exists a valid model $T$ such
that $T(u) = t$.

If we compare this new interpretation of G-ACGs to the former one, we find out
that in the cases when the graph of the grammar is an arborescence, they are
actually equivalent. This means that in classical ACGs, where all the diagrams
are arborescences, the two metaphors (nodes as languages and nodes as terms)
can be, and are, used interchangeably. However, as we start to consider
non-arborescent graphs, we find, interestingly, that the two paradigms diverge
(i.e. $\exists \mathcal{G}, u.\ \mathcal{E}_{\mathcal{G}}(u) \neq
\mathcal{P}_{\mathcal{G}}(u)$).

@@

We start with the extraction constraints. \cite{pogodalla2012controlling}
describes three different constraints on extraction, all of them expressed
using the Constraint pattern (Figure~\ref{fig:single-cont}). We can take the
union of these three G-ACGs to get the G-ACG on Figure~\ref{fig:merged-conts}.
Next, we can incorporate knowledge from \cite{pogodalla2007generalizing}, by
splitting the $\mathcal{L}_{Syn}$ lexicon and introducing a new intermediate
signature (Figure~\ref{fig:conts-elab}). The new signature, $Syntax$, deals
purely with syntax without any issues of operator scope and its functions have
types of a lower order than those in $Syn$. We will finally add our
contribution, the signatures $Neg$ and $Agr$ and the lexicons
$\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ which implement the constraint on
negation we discussed before and some simple notion of agreement, respectively
(Figure~\ref{fig:final-synth}).

@@

We have decomposed our grammar into 9 signatures and 8 lexicons. Besides
making the grammar more readable and easily navigable, it also makes the
grammar more attractive from a computational point of view. The complexity of
parsing the highest abstract-most structures and verifying all the constraints
is decomposed into smaller subtasks. A string can be parsed into its
representation at the $Syntax$ level fairly easily (compared to parsing it at
the higher levels) since $Syntax$ employs types of lower orders than the more
abstract signatures. Finding the syntactic structure for a string is thus
feasible. We then proceed to verifying the syntactic constraints. Here we can
observe that the lexicons $\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ are mere
relabelings and therefore parsing (i.e. constraint verification) is trivial
(i.e. decidable). Now in order to determine the meaning of our sentence, we
have to continue parsing by finding an antecedent in $Syn$. This involves
higher-order types used to encode the various scope-bearing
operators. Finally, we can opt in to verify the high-level syntax-semantics
constraints which employ dependent types
\cite{pogodalla2012controlling}. However, the lexicons in this case resemble
relabelings and a semi-decidable procedure could be used to solve the
problem. Having the grammar decomposed like this opens it up for computational
processing by letting us engage in different tactics at different stages
instead of having to parse a single hidden structure which enforces
grammaticality w.r.t. every constraint. G-ACGs share these perks also with
classical ACGs but they let use this composition in another dimension, which
facilitates the writing of multiple constraints.
